---
title: 'Geog4/6300: Lab 4'
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## One and two sample difference tests

**Overview:**
In this lab, you'll be doing analysis with public use microdata from the Household Pulse survey administered by the U.S. Census Bureau during the COVID pandemic. The survey had a particular focus on pandemic effects on health and wellbeing, but it also included questions on the impact of natural disasters on households. That's the section you'll be focusing on here. 

Unlike the census data you're probably used to, microdata provides anonymized individual level responses that researchers can use to calculate their own aggregate statistics.

There's more info on the Pulse survey and the public use microdata file at this link:

* https://www.census.gov/programs-surveys/household-pulse-survey/data/datasets.html

The main data are contained in the `pulse_2023` csv file in the data folder. The data dictionary for this dataset is also available in the data folder.

You'll use this data to answer two research questions:

* Did households displaced by natural disasters include more children than those not displaced? 
* Is there a relationship between household income and a household's ability to return to its place of residence after being displaced? 

## Assessment criteria
For the lab to be marked as complete, the following criteria must be met:

* Criteria 1: Your responses will show an ability to filter survey data for appropriate responses. (Tasks 1 and 7)
* Criteria 2: Your responses will demonstrate that you can test a variable for normality and decide on the appropriate statistical test. (Tasks 2, 3, and 4)
* Criteria 3: Your responses will show your ability to correctly use two-sample tests of difference and chi-square tests. (Tasks 5, 6, 9, and 10)
* Criteria 4: Your responses will accurately and descriptively interpret the results of your statistical tests and explain their significance. (Tasks 7 and 11)

###Part 1: Loading/filtering data and checking for normality.

In the first part of the lab, you'll be using two variables: THHLD_NUMKID (number of children in a household) and ND_DISPLACE (whether a household was displaced by a natural disaster in the last year). 

Load the dataset using read_csv, and also take a look at the two variables of interest in the data dictionary. For this lab, you only want observations with *valid responses*--filtering out any where there were no answers. Filter these data so you have only valid responses. 

**Task 1** _Load the Pulse data and filter it to only valid responses._

```{r Task1}
#Code here

```

You'll be doing a two sample test of difference to look at whether the number of children differs between those displaced and not displaced. To know which test to use, you have to assess the normality of the THHLD_NUMKID variable. 

To do so, create a random sample of 5,000 records using the `slice_sample` function (look it up in help for more information). That's because the Shapiro-Wilk test can't handle more than 5,000 records. Create that sample using `slice_sample` and then do the following: 1) Perform a Shapiro-Wilk normality test on this variable. 2) Create a Q-Q plot with a line showing the normal distribution. 3) Create a histogram using ggplot or the base R function.

**Task 2** _Create a random 5,000 record sample from the Pulse dataset. Then run a Shapiro-Wilk normality test, create a QQ plot with a line showing normality, and create a histogram._

```{r Task2}
#Code here.

```


**Task2 3** _Based on the results you saw in question 2, is the THHLD_NUMKID variable normally distributed? How do each of those three tests inform your decision?_

{Your response here}

###Part 2: Testing the difference in number of children.###

Now you can test whether the number of children is different in between displaced and non-displaced households.

**Task 4** _Based on the normality of the data, what is the appropriate statistical test for determining if the actual number of children differs between these two groups? Explain why this test is appropriate and give a null and alternative hypothesis for this test_

{Your response here.}

**Task 5** _Run the test you identified in question 4. Be sure to use the FULL sample for this test (not the smaller sample from Question 2). Also calculate the mean for each dataset, which you can use to determine the nature of the difference--which group (if any) has more children._

```{r Task5}
#Code goes here
```

**Task 6** _Interpret the results of the analysis you did in task 5. Do you reject the null hypothesis? Why or why not? What can we say about any difference in the number of children in these two groups, focusing on the significance of the test result and the difference in means within this sample?_

{Your response here.}

###Part 3: Identifying differences in time displaced by natural disasters across income groups

In Part 2, you used a test of difference that included a continuous and categorical variable. In this section, you'll focus only on households displaced by natural disasters to identify statistically significant differences in time displaced using two categorical variables: one placing responses into groups by income (INCOME) and the other showing the time displaced (ND_HOWLONG). Again, look at the data dictionary to better understand how these two variables are structured.

**Task 7** _Reload the original, unfiltered Pulse dataset (not the one you used in parts 1 and 2). Select only observations for households displaced by a natural disaster using the ND_DISPLACE variable. Then filter for records with valid responses to INCOME and ND_HOWLONG. Call the head of the table with `kable` when done_

```{r Task7}
#Code goes here.
```

**Task 8** _Create a contingency table of responses to the ND_HOWLONG variable by INCOME. The rows of this table should be the income groups and the columns should be the categorical variables showing time displaced. Call the table when you are done._

```{r Task8}
#Code goes here.
```

**Task 9** _Identify if there are statistically significant differences in displacement time between these income groups by conducting a chi square test on the contingency table in task 8. You'll likely need to delete the INCOME variable to do so._ 

```{r Task9}
#Code goes here
```

**Task 10** _To help with your interpretation of results, convert the contingency table in Task 8 from one that shows the number of responses to one that shows the percentage of responses within each income group. So in each row, the number should show the percentage of responses in that income group that selected that response for ND_HOWLONG._

```{r Task10}
#Your code goes here
```

**Question 11** _Interpret the results of your analysis in tasks 9 and 10. What were your null and alternative hypotheses for the chi square test? Do you reject the null hypothesis based on these results and why (or why not)? What does the table in task 9 tell you about any differences in displacement across income groups?_

{Response goes here.}

###Challenge question:
Pick a variable of interest from the pulse dataset and create an appropriate state level estimate for it using the EST_ST variable, which is the state FIPS code. For example, you could look at the percentage of the population displaced (ND_DISPLACE) by state. Or you could map where displacement by flooding was most common using the ND_TYPE2 variable. 

Once you've calculated that estimate, map it out using the `states_albers` geopackage file in the data folder for this lab. 

```{r}
#Code goes here
```

### Final submission stuff

**Disclosure of assistance:** Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?

{Response here.}

**Lab reflection:** How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What are the biggest things you learned by completing it?

{Your response here}
